Relationships in the extremes and their influence on competition and coexistence

### 2021-09-22 paper/Paper.Rnw
editing lit review para-mean shifts

The consequence of climate change that has the most potential to influence ATAs in the environment the increase in intensity, duration, extent, and frequency of extreme climatic events \citep{Meehl_2004, Ummenhofer_2017}, i.e., heat-waves, floods, hurricanes, fires, etc. 

@@

%***JAS: I am trying to relay the message that climate change and specifically extreme events have consequences for coexistence and competition
% Maybe a better way to start this paragraph is:
% EC covariance, and thus the storage effect, may be altered by changes in the environment. 
% need an indication that we are more interested in increased variance and extreme events because we are looking we are looking for changes in a co-varying relationship??
%***DAN: Possible topic sentence like thing after talking to Jasmin on 9/17: 
%Although, to our knowledge, the potential importance of climatic changes in ATAs on coexistence has not been studied,
%the importance for coexistence of other aspects of climate change, including changes in means and variability of environmental variables, 
%have been studied in several systems. 
%***DAN: Do it like you have it but shorter.
Although, to our knowledge, the potential importance of climatic changes in ATAs on coexistence has not been studied, the importance for coexistence of other aspects of climate change, including changes in means and variability of environmental variables, have been studied in several systems.  
Environmental fluctuations are directly impacted by climate change through changes in means and increased variance, both of which can be attributed to extreme events \citep{Hansen_2012}.
Currently, effects on competition and coexistence due to climate change are more widely researched through studying changes in the mean of environmental variables. 
Increased variance is less studied. 
% thus less likely to be a result of extreme events. 
%influenced by extreme climatic events as they increase in frequency, intensity, duration, and extent \citep{Meehl_2004, Ummenhofer_2017}. 
%Thus, the rampage of extreme events such as heat-waves, floods, hurricanes, fires, etc. implications for coexistence and competition. 
%Jasmin tried to split this para up into effects on coexistence of changes in mean, starting here... 
Changes in means can be attributed to extreme events; however changes in means are more-so a result of gradual shifts over time, whereas increased variance is more likely a product of extreme events as they push environmental variables to their extremes.

For example, decreased mean growth season precipitation can decrease the competitive ability of dominant forest tree species \citep{Clark_2011}. This may be detrimental if the dominant species is crucial to the ecosystem or it may not be as rare species can increase in diversity while dominant species become excluded \citep{Mariotte_2013}. Conversely, changes in environmental means can increase fitness of a dominant species and in turn reduce community diversity, typical of invasive species \citep{Vitousek_1997}. \cite{Descombes_2020}, through experimental manipulation, found that range shifts, as a product of climate change, favored coexistence between alpine plants while \cite{Usinowicz_2021} predict increases in competitive pressures as the climate change induced range shifts lead populations to interact with new conspecifics. Similarly, \cite{diPaola_2012} predict the collapse of coexistence between Mediterranean tree species as mean rainfall levels are forecasted to decrease.    
%...and variance...
In contrast, increased environmental variance has been less commonly studied; however, more research on the effect of increased variance has been urged and received recent focus through the study of extreme climatic events \citep{Jentsch_2007, Smith_2011}. Extreme drought and heat waves impact competitive output and community composition of grassland plants\citep{White_2000, White_2001, Hoover_2014}. However, complete exclusion of dominant species due to extreme events is uncommon \citep{Lloret_2012}. Moreover, in regard to invasive species, increased climate and resource variability allowed for greater success of invasive plants in terms of density and competitive equilibrium \citep{Parepa_2013}. 
%...and extreme events (though this blends together with changes in variance)
Extreme events constitute large disturbances, quantitatively amounting to large fluctuations in environmental drivers and their ecological responses \citep{Smith_2011}. Although, increased variability of environmental fluctuations is a newer consideration for studying environmental changes; environmental variability has been analytically shown to be important for classical coexistence \citep{Chesson_1981, Warner_1985}. Moreover, increased variability has been found to promote coexistence in plants communities \citep{Adler_2006,  Angert_2009}, but disrupt coexistence in birds \citep{Stenseth_2002, Stenseth_2015}. This present study has applications to extreme events and climate change as we consider relationships in their extremes, potentially caused by extreme events, and their implications for competition and coexistence.
%DAN: So what jasmin was trying to do was summarize a lot of lit on effects of climate change on coexistence, in these three parts
%(means, varinace, and extreme events).

>>>>

The consequence of climate change that has the most potential to influence ATAs in the environment the increase in intensity, duration, extent, and frequency of extreme climatic events \citep{Jentsch_2007, Meehl_2004, Ummenhofer_2017}, i.e., heat-waves, floods, hurricanes, fires, etc. 

@@

%***JAS: I am trying to relay the message that climate change and specifically extreme events have consequences for coexistence and competition
% Maybe a better way to start this paragraph is:
% EC covariance, and thus the storage effect, may be altered by changes in the environment. 
% need an indication that we are more interested in increased variance and extreme events because we are looking we are looking for changes in a co-varying relationship??
%***DAN: Possible topic sentence like thing after talking to Jasmin on 9/17: 
%Although, to our knowledge, the potential importance of climatic changes in ATAs on coexistence has not been studied,
%the importance for coexistence of other aspects of climate change, including changes in means and variability of environmental variables, 
%have been studied in several systems. 
%***DAN: Do it like you have it but shorter.
%the RAMPAGE of extreme events such as heat-waves, floods, hurricanes, fires, etc. implications for coexistence and competition. 
Although, to our knowledge, the potential importance of climatic changes in ATAs on coexistence has not been studied, the importance for coexistence of other aspects of climate change, including changes in means and variability of environmental variables, have been studied in several systems.  
Environmental fluctuations are directly impacted by climate change through changes in means and increased variance \citep{Hansen_2012}. While both changes can be attributed to extreme events, shifts in mean are more-so a result of gradual trend effects, and increased variance is a result of extreme event effects \citep{Jentsch_2007} .
Impacts on competition and coexistence of climate change have been more commonly researched via studying changes in the mean of environmental variables; impacts of increased variance is less studied, although the rate of study has increased. \citep{Jentsch_2007}  %importance has been realized and variability is becoming more commonly studied now

%Jasmin tried to split this para up into effects on coexistence of changes in mean, starting here...
For example, decreased mean growth season precipitation can decrease the competitive ability of dominant forest tree species \citep{Clark_2011}, and as such, \cite{diPaola_2012} predict the collapse of coexistence between Mediterranean tree species as mean rainfall levels are forecasted to decrease. 
On the contrary, dominant species exclusion may be beneficial as rare species can increase in diversity\citep{Mariotte_2013}. Changes in environmental means can also increase fitness of a dominant species and in turn reduce community diversity, typical of invasive species \citep{Vitousek_1997}. 
As a result of changing means, range shifts can favor plant coexistence by introducing new herbivores\cite{Descombes_2020} or disfavor coexistence through increased competition \cite{Usinowicz_2021}.  
%...and variance...
In contrast, increased environmental variance has been less commonly studied; however, more research on the effect of increased variance has been urged and received recent focus through the study of extreme climatic events \citep{Jentsch_2007, Smith_2011}. Extreme drought and heat waves impact competitive output and community composition of grassland plants\citep{White_2000, White_2001, Hoover_2014}. However, complete exclusion of dominant species due to extreme events is uncommon \citep{Lloret_2012}. \cite{Fay_2003} experimentally manipulated rainfall for more frequent and intense precipitation while keeping constant the total volume of rain over Kansas tallgrass prairie and found that while some species had reduced fitness, overall community structure remained the same. Moreover, in regard to invasive species, increased climate and resource variability allowed for greater success of invasive plants in terms of density and competitive equilibrium \citep{Parepa_2013}. 
%...and extreme events (though this blends together with changes in variance)
Extreme events constitute large disturbances, quantitatively amounting to large fluctuations in environmental drivers and their ecological responses \citep{Smith_2011}. Although, increased variability of environmental fluctuations is a newer consideration for studying environmental changes; environmental variability has been analytically shown to be important for classical coexistence \citep{Chesson_1981, Warner_1985}. Moreover, increased variability has been found to promote coexistence in plants communities \citep{Adler_2006,  Angert_2009}, but disrupt coexistence in birds \citep{Stenseth_2002, Stenseth_2015}. This present study has applications to extreme events and climate change as we consider relationships in their extremes, potentially caused by extreme events, and their implications for competition and coexistence.
%DAN: So what jasmin was trying to do was summarize a lot of lit on effects of climate change on coexistence, in these three parts
%(means, varinace, and extreme events).
